VariantSpark: Population Scale Clustering of Genotype Information

### 2015-08-18 doc/publications/journal2014/paper.tex
final section 1

In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark\ against ADAM as well as the more traditional approaches using Hadoop Mahout, Python and R. 
We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption rendering a whole-genome input infeasible.  
Furthermore, we perform the comparisons on a single virtual-machine to ensure the five different approaches have access to the same resources.
We use k-means clustering algorithm in the respective implementations, which all require the VCF input files to be pre-processed (see methods). 
We therefore compare the time required for the pre-processing step, as well as the time required for k-means clustering.
% DENIS: include this in the table too
%The python and R pre-processing implementation only runs on a single thread, so we don't include this in the comparison.

%We compare the time required to cluster individuals based on their chromosome 22 variants using various implementations. We compare our Variant{\sc Spark} implementation with ADAM and our Hadoop implementation, as well as a Python and R implementation.
%To keep the comparison fair for the Python and R implementations, we perform the comparisons on a single virtual-machine, rather than a cluster. Therefore, for k-means clustering, each implementation has access to the same resources.
%Each of the k-means clustering implementations that we compare require the input to be row vectors. We need to pre-process the VCF input files to this format, so we compare the time required for the pre-processing step, as well as the time required for k-means clustering.
%The python and R pre-processing implementation only runs on a single thread, so we don't include this in the comparison.

As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark\, requiring 2mins and 58secs. %178s 
This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
or our Hadoop implementation(14min and 22secs). %862s
Unlike \variantSpark\ or our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format. 
VCF files must hence be converted to this binary format before they can be pre-processed into the row vector format. 
Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs %793s
rendering our approach an order of magnitude faster. 
[Results for Python and R here].
This substantially different runtime for the pre-processing in Python and R is due to these implementations does not natively supporting multithreading while \variantSpark, ADAM and Hadoop using XX CPUs.

%Pre-processing the data is fastest in \variantSpark\, requiring 2mins and 58secs. Next fastest is the ADAM implementation, requiring 12min and 48secs. 
%Our Hadoop implementation took 14min and 22secs.
%Unlike Variant{\sc Spark} and our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format. VCF files must be converted to this binary format before they can be pre-processed into the row vector format. Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs.

Clustering the samples 30\% fastest in \variantSpark\ compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
The similarity between \variantSpark\ and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
The 30 second speed increase in \variantSpark\ is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors.
%Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
We also compared the k-means clustering times to a multi-threaded approach in Python and R.

Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation. Python required 11min and 29sec, whereas R required 7min and 25sec. 
This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on): the cost of disk IO. Where Hadoop's advantage is that it can scale to datasets that exceed memory size, in this example, the entire dataset can fit in memory on a single machine.
So, where the k-means implemtations we use in R and Python store the output from each k-means iteration in memory, Hadoop writes the output of each iteration to disk, whether or not the amount of available memory is sufficient to store this data.

This unnecessary disk IO forms a bottleneck that results in the longer run-times. {\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.

%Clustering the samples is also fastest in \variantSpark\, albeit by a smaller margin. The time required is 1min and 20secs compared to 1min and 52secs, in Variant{\sc Spark} and ADAM, respectively. Hadoop however, required 14mins and 23secs.
%The similarity between Variant{\sc Spark} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. The 30 second speed increase in Variant{\sc Spark} is likely due to Variant{\sc Spark}'s pre-processing implementation converting VCF files to sparse vectors, whereas ADAM creates dense vectors.
%Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.

%We also compared the k-means clustering times to a multi-threaded approach in Python and R. Although slower than the {\sc Spark} implementations in Variant{\sc Spark} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation. Python required 11min and 29sec, whereas R required 7min and 25sec. This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on). Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster. However, in this example, the entire dataset can fit in the memory on a single machine. The k-means implemtations
%we use in R and Python cluster the in-memory row vectors. Hadoop, however, writes the output of each MapReduce iteration to disk. This unnecessary disk IO forms a bottleneck that results in the longer run-times. {\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.

%To summarise the total times for each big-data platform, Variant{\sc Spark} was 4mins and 18secs, ADAM was 14mins and 14secs (or 27mins and 53secs including the one-off conversion of VCF to ADAM) and our Hadoop implementation was 28mins and 45secs.

@@

To demonstrate the scalability of {\sc Spark}, we continue to cluster the same individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
Pre-processing the data takes approximately 12 minutes for a 20\% sample. For 40\% of the genome, the time required is about 19 minutes, and finally, to preprocess 100\% of the genome, the time required is about 40 minutes.
For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for the entire autosome. Memory requirements for clustering, however, increase by a

>>>>

In this section we compare the time required to cluster individuals based on genomic variants using \variantSpark\ against ADAM as well as the more traditional approaches using Hadoop Mahout, Python and R. 
We limit the genomic variants to only chromosome 22 as the traditional approaches have substantially larger memory consumption rendering a whole-genome input infeasible.  
Furthermore, we perform the comparisons on a single virtual-machine to ensure the five different approaches have access to the same resources.
We use k-means clustering algorithm in the respective implementations, which all require the VCF input files to be pre-processed (see methods). 
We therefore compare the time required for the pre-processing step, as well as the time required for k-means clustering.

As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark\, requiring 2mins and 58secs. %178s 
This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
or our Hadoop implementation(14min and 22secs). %862s
Unlike \variantSpark\ or our Hadoop implementation, which pre-process VCF files directly, ADAM requires the input to be in their ADAM file format. 
VCF files must hence be converted to this binary format before they can be pre-processed into the row vector format. 
Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs %793s
rendering our approach almost an order of magnitude faster. 
R on the other hand takes 67 minutes and Python 92 minutes.
This substantially different runtime for the pre-processing is due to the Python and R implementations not natively supporting multithreading while \variantSpark, ADAM and Hadoop using 8 CPUs.

Clustering the samples 30\% fastest in \variantSpark\ compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
The similarity between \variantSpark\ and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
The 30 second speed increase in \variantSpark\ is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors making operations less efficient.

We also compared the k-means clustering times to a multi-threaded approach in Python and R. 
Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation; Python required 11min and 29sec, whereas R required 7min and 25sec. 
This is due the dataset (chromosome 22) being small enough to fit into memory on a single machine, which enables R and Python to store the output from each k-means iteration directly in memory whereas Hadoop is designed to write the output of each iteration to disk. 
This demonstrates the improvements of {\sc Spark} over Hadoop where datasets can be processed in memory, virtually eliminating the disk IO bottleneck on smaller datasets.

To summarise, \variantSpark\ completes the whole variant clustering task in just over 4 minutes, which is 80\% faster than ADAM (28 mins), 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).

@@

To demonstrate the scalability of {\sc Spark}, we continue to cluster the same individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
Pre-processing the data takes approximately 12 minutes for 20\% of the genome, 19 minutes for 40\% and 40 minutes for 100\% of the genome, respectively.
For each of these sample sizes, the memory required doesn't exceed a modest 2GB per executor, even for the entire autosome. Memory requirements for clustering, however, increase by a
